% ============================================================
%  ECON306: W09D3 — Is QALY-Max Enough? MCDA Concepts and Methods
%          (PAPRIKA, 1000minds); Discrete Choice Experiments;
%          ISPOR Good Practice
%  Date: Friday 18 September 2026
%  SOURCE: COMPRESSION of old w11_week11 (MCDA sections) from FULL deck
%    (lines 3382–3567). Best frames per Course Map W09D3 LOs.
%  DROPPED from old deck: Week 11 Overview frame (week numbering wrong),
%    "Ageing populations and MCDA" frame (content folded into MCDA good
%    practice frame here), duplicate PAPRIKA description. All logged.
% ============================================================

\documentclass[aspectratio=169, 12pt]{beamer}

\input{../../Preambles/econ306_preamble.tex}

\title[ECON306: W09D3 — MCDA and DCEs]{ECON306\\The Economics of Health and Education}
\subtitle{W09D3: Is QALY-Max Enough? MCDA Concepts and Methods; Discrete Choice Experiments}
\author{Austin Henderson}
\institute{Department of Economics\\University of Otago}
\date{Friday 18 September 2026}

\begin{document}

\begin{frame}[plain]
  \titlepage
\end{frame}

\recapslide{
  \item QALY maximisation is efficient but may violate equity --- who benefits matters
  \item Distributional weights ($w_i$) and social welfare functions formalise equity concerns
  \item League tables ignore equity, budget impact, uncertainty --- PHARMAC uses multi-criteria process
}

\planslide{Beyond QALYs: MCDA, DCEs, and Decision Support}{
  \item Is QALY maximisation enough? Limitations recap
  \item What is MCDA and why does it matter for health economics?
  \item A brief history of decision analysis
  \item The PAPRIKA method and 1000Minds (Otago)
  \item Discrete choice experiments in health
  \item ISPOR good-practice framework
}

\begin{frame}{What is MCDA?}
  \small
  A family of methods for \textbf{structuring and solving complex decisions} involving multiple,
  often conflicting, criteria --- where no single option is best on all criteria.
  \vspace{0.2em}
  \begin{columns}
    \column{0.52\textwidth}
      \small Health allocation decisions \emph{always} involve multiple criteria:
      \begin{itemize}
        \item Clinical benefit (QALYs)
        \item Equity (who benefits)
        \item Cost and budget impact
        \item Severity and unmet need
        \item Quality of evidence
        \item Innovation incentives
      \end{itemize}
    \column{0.45\textwidth}
      \begin{block}{Why MCDA?}
        \small CUA captures only some of these. MCDA \textbf{explicitly structures all criteria} and applies weights --- making trade-offs transparent, not hidden in committee deliberations.
      \end{block}
  \end{columns}
\end{frame}

\begin{frame}{A Brief History of Decision Analysis}
  \begin{itemize}
    \item \textbf{1950s--60s}: Operations research and statistical decision theory
          (Wald, Savage); decision matrices
    \item \textbf{1970s}: Multi-attribute utility theory (MAUT): Keeney \& Raiffa (1976)
    \item \textbf{1980s}: Analytic Hierarchy Process (AHP): Saaty (1980) --- widely used
          but criticised for rank reversal
    \item \textbf{1990s--2000s}: MCDA applied to environmental and public policy;
          beginning of health applications
    \item \textbf{2010s}: ISPOR MCDA Emerging Good Practices Task Force (2016);
          rapid adoption in health technology assessment globally
    \item \textbf{NZ contribution}: \textbf{Paul Hansen \& Franz Ombler (2008)}: developed
          the PAPRIKA method, implemented in \textbf{1000Minds} software; now used by
          NZ Ministry of Health, WHO, and hundreds of organisations globally
  \end{itemize}
\end{frame}

\begin{frame}{The PAPRIKA Method and 1000Minds}
  \small
  \textbf{PAPRIKA} (Hansen \& Ombler, 2008): Potentially All Pairwise Rankings of All possible
  Alternatives. Elicits weights through \textbf{pairwise ranking of partial profiles} --- no
  numerical weight assignment required.
  \vspace{0.2em}
  \begin{columns}
    \column{0.55\textwidth}
      \footnotesize
      \begin{itemize}
        \item \textbf{1000Minds software}: NZ Ministry of Health ($>$100,000 patients/year),
              WHO, health ministries in Australia, UK, and elsewhere
        \item \textbf{Advantages}: avoids anchoring bias, handles many criteria, produces
              consistent weights, user-friendly
        \item \textbf{Tutorial 6 (Week 12)}: use 1000Minds for a life-advice MCDM exercise
              (``what do you value, what to do, where to live'')
      \end{itemize}
    \column{0.42\textwidth}
      \begin{exampleblock}{Golan \& Hansen (2012)}
        \footnotesize Apply PAPRIKA/1000Minds to Israeli pharmaceutical prioritisation --- a worked
        example of MCDA in practice. Required reading for this week.
      \end{exampleblock}
  \end{columns}
\end{frame}

\begin{frame}{Discrete Choice Experiments in Health Economics}
  \begin{itemize}
    \item \textbf{Discrete Choice Experiments (DCEs)}: present respondents with hypothetical
          choices between alternatives described by attributes; infer preferences from choices
    \item Applications in health:
    \begin{itemize}
      \item Valuing health-related quality of life (generating EQ-5D value sets)
      \item Understanding patient preferences for health care features
      \item Valuing GP attributes (waiting time, doctor continuity, after-hours access)
      \item Measuring public preferences for health system attributes
    \end{itemize}
    \item \textbf{Advantage over surveys}: reveals preferences through choices (analogous
          to revealed preference), not just stated intentions
    \item \textbf{NZ application}: Sullivan \& Hansen (2017) used DCEs to determine
          public preferences for health technology prioritisation criteria in NZ
  \end{itemize}
\end{frame}

\begin{frame}{MCDA Good Practice: The ISPOR Framework}
  \begin{columns}
    \column{0.58\textwidth}
      \small \textbf{ISPOR MCDA Emerging Good Practices (2016):}
      \footnotesize
      \begin{enumerate}
        \item Define the decision problem
        \item Select and structure criteria
        \item Measure performance on each criterion
        \item Score the options
        \item Weight the criteria
        \item Calculate an overall value score
        \item Examine results for robustness (sensitivity analysis)
        \item Use results to inform (not replace) decision-making
      \end{enumerate}
    \column{0.38\textwidth}
      \begin{alertblock}{Key principle}
        \footnotesize MCDA is a \textbf{decision support} tool, not a decision rule. It informs and
        structures deliberation; it does not replace human judgement.
      \end{alertblock}
  \end{columns}
\end{frame}

\begin{frame}{MCDA and Ageing Populations}
  \begin{columns}
    \column{0.58\textwidth}
      \footnotesize As NZ's population ages, MCDA frameworks face new pressures:
      \begin{itemize}
        \item How should \textbf{long-term care quality} be weighted relative to acute treatment?
        \item Should \textbf{caregiver burden} enter as a criterion?
        \item How do we weight \textbf{dementia care} when cognitive impairment limits
              self-reported quality of life?
        \item Should there be an \textbf{``ageing premium''} analogous to the UK end-of-life premium?
      \end{itemize}
      \vspace{0.15em}
      \footnotesize \textbf{1000Minds} is already used by NZ aged care providers for needs assessment and prioritisation.
    \column{0.38\textwidth}
      \begin{block}{Assignment reminder}
        \small Due \textbf{Mon 5 Oct 2026, 9:00 am}.\\[0.2em]
        Apply MCDA reasoning in your PHARMAC funding brief.\\[0.15em]
        Full details on Brightspace (Aoroa).
      \end{block}
  \end{columns}
\end{frame}

\summaryside{
  \item MCDA explicitly structures and weights multiple criteria; more transparent than ad-hoc committee deliberation
  \item PAPRIKA/1000Minds (Hansen \& Ombler, 2008): world-leading MCDA tool developed at Otago, used by NZ Ministry of Health and WHO
  \item DCEs reveal preferences for health system attributes through hypothetical choices; used in NZ for prioritisation criteria (Sullivan \& Hansen, 2017)
  \item ISPOR framework: 8 steps for MCDA in health; it supports rather than replaces decision-making; Tutorial 6 applies 1000Minds to life choices
}

\end{document}
